Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-05-04 causal-bandit/nips_2016/general.tex
adds clarifications

We now consider the more general problem where the graph structure is known, but arbitrary. For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation). However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \cite{Pearl2000}. 
\eq{
\P{X_1...X_N|do(X_i=j)} = 
\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)  
} 
Where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.

@@

To address this difficult problem, we firstly note that the causal effect of an action $a$ on $Y$ can be factored into the causal effect of $a$ on the parents of $Y$ and the effect of the parents of $Y$ on $Y$. $\P{Y|a} = \P{Y|\parents{Y}}\P{\parents{Y}|a}$ We then assume the conditional interventional distributions $\P{\parents{Y}|a}$ are known. ADD MORE JUSTIFICATION FOR THIS ASSUMPTION. IE COULD HAPPEN IF WE HAD EXPERIMENTAL DATA COLLECTED WITH A DIFFERENT OBJECTIVE AND IS A NESSESSARY FIRST STEP TO SOLVING THE MORE GENERAL PROBLEM.

@@

Our algorithm will choose $T$ samples from $Q$ IS THIS LANGUAGE CORRECT. CAN WE SAY IT LIKE THIS? and use them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via a truncated importance weighted estimator. The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail that is so detrimental to its concentration guarantees.

Define the random variable
\eq{
Z_a(X) = R_a(X) Y \ind{R_a(X) \leq B_a}\,,
}
where $B_a \geq 0$ is some constant to be chosen subsequently and
\eq{
R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\,.
}
Abusing notation we define $m(\eta)$ by
\eq{
m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
}
We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately 
coincides with the version for parallel bandits, justifying the name overloading.

\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $a_t = a$ with probability $\eta$
\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\STATE For each $a \in \calA$ compute an estimate of its return:
\eq{
\forall a\in \calA \qquad \hat \mu_a = \frac{1}{T} \sum_{t=1}^T Z_a(X_t)
}
\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}

>>>>

We now consider the more general problem where the graph structure is known, but arbitrary.

% How do actions map to interventions (we need not only consider singleton interventions).
%

For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation). However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \cite{Pearl2000}. 
\eq{
\P{X_1...X_N|do(X_i=j)} = 
\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)  
} 
Where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.

@@

To address this difficult problem, we firstly note that the causal effect of an action $a$ on $Y$ can be factored into the causal effect of $a$ on the parents of $Y$ and the effect of the parents of $Y$ on $Y$. $\P{Y|a} = \P{Y|\parents{Y}}\P{\parents{Y}|a}$. We then assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) are known. These could be estimated from experimental data on the same covariates $\boldsymbol{X}$, but where the outcome of interest differed such that $Y$ was not included, or similarly from observational data subject to identifiability constraints.

@@

\begin{wrapfigure}[14]{r}{0.5\textwidth}
\vspace{-12pt}
\begin{minipage}{.5\textwidth}
\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $a_t = a$ with probability $\eta$
\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\STATE For each $a \in \calA$ compute an estimate of its return:
\eq{
\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
}
\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}
\end{minipage}
\end{wrapfigure}

Our algorithm samples $T$ actions from $\eta$ and uses them to estimate the returns $\mu_a$ for all $a \in \calA$ simultaneously via a truncated importance weighted estimator. Let $\parents{Y}(X)$ denote the realization of the variables in $X$ that are parents of Y and define $R_a(X) = \frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}$

\eq {
\hat{\mu_a} = \frac{1}{T} \sum_{t=1}^T R_a(X) Y \ind{R_a(X) \leq B_a}\,, 
}

where $ B_a \geq 0$  is a constant that tunes the level of truncation to be chosen subsequently. The truncation introduces a bias in the estimator, but simultaneously chops the potentially heavy tail that is so detrimental to its concentration guarantees.

The distribution over actions, $\eta$ plays the role of allocating samples to actions and is optimized to minimize the worst-case simple regret. Abusing notation we define $m(\eta)$ by
\eq{
m(\eta) = \max_{a \in \calA} \EE_{\Ps_a}\left[\frac{\Pn{a}{\parents{Y}(X)}}{\Q{\parents{Y}(X)}}\right]\,.
}
We will shortly see that $m(\eta)$ is a measure of the difficulty of the problem that approximately coincides with the version for parallel bandits, justifying the name overloading.

### 2016-05-04 causal-bandit/nips_2016/setup.tex
adds clarifications

We now introduce a novel class of stochastic sequential decision problems which we call \emph{causal bandit problems}. 
In these problems, rewards are given for repeated interventions on a fixed causal model~\cite{Pearl2000}. 
Following the terminology and notation in~\cite{Koller2009}, a \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X} = \{ X_1, \ldots, X_N \}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$.
We will assume each variable only takes on a finite number of distinct values.
An edge from variable $X_i$ to $X_j$ is interpreted to mean that a change in the value of $X_i$ may directly cause a change to the value of $X_j$.
The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$ or just $\parents{i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
The intervention also ``mutilates'' the graph $\mathcal{G}$ by removing all edges from $\parents{i}$ to $X_i$ for each $X_i \in \vec{X}$. 
The resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$. 
Details can be found in Chapter 21 of~\cite{Koller2009}.

>>>>

We now introduce a novel class of stochastic sequential decision problems which we call \emph{causal bandit problems}. 
In these problems, rewards are given for repeated interventions on a fixed causal model~\cite{Pearl2000}. 
Following the terminology and notation in~\cite{Koller2009}, a \emph{causal model} is given by a directed acyclic graph $\mathcal{G}$ over a set of random variables $\mathcal{X} = \{ X_1, \ldots, X_N \}$ and a joint distribution $\mathrm{P}$ over $\mathcal{X}$ that factorises over $\mathcal{G}$.
We will assume each variable only takes on a finite number of distinct values.
An edge from variable $X_i$ to $X_j$ is interpreted to mean that a change in the value of $X_i$ may directly cause a change to the value of $X_j$.
The \emph{parents} of a variable $X_i$, denoted $\parents{X_i}$, is the set of all variables $X_j$ such that there is an edge from $X_j$ to $X_i$ in $\mathcal{G}$.
An \emph{intervention (of size $n$)}, denoted $do(\vec{X}=\vec{x})$, assigns the values $\vec{x}=\{x_1, \ldots, x_n\}$ to the corresponding variables $\vec{X}=\{X_1, \ldots, X_n\} \subset \mathcal{X}$ with the empty intervention (where no variable is set) denoted $do()$.
The intervention also ``mutilates'' the graph $\mathcal{G}$ by removing all edges from $\parents{i}$ to $X_i$ for each $X_i \in \vec{X}$. 
The resulting graph defines a probability distribution $\P{\vec{X}^c | do(\vec{X}=\vec{x})}$ over $\vec{X}^c := \mathcal{X} - \vec{X}$. 
Details can be found in Chapter 21 of~\cite{Koller2009}.

### 2016-05-04 causal-bandit/nips_2016/simple-regret.tex
adds clarifications

This captures a trade off between identifying the low probability variables and not having too many of them.
When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 1$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$, the number of variables.

>>>>

$I_\tau$ represents the set of actions defined as low probability given a threshold $\tau$. As $\tau$ is increased,

This captures a trade off between identifying the low probability actions and not having too many of them. $I_\tau$ represents the actions defined as 'low p
When $\vec{q} = (\frac{1}{2}, \ldots, \frac{1}{2})$ we have $m(\vec{q}) = 1$ and when $\vec{q} = (0, \ldots, 0)$ we have $m(\vec{q}) = N$, the number of variables.
